---
title: "Stats 102B Homework 2"
author: "Christian Jordan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r message = FALSE, warning = FALSE, echo = FALSE}

```

# (1) Two-Dimensional Random Search

## (a) random_search()
```{r}
random_search <- function(g, alpha_choice = 1, max_its = 100, num_samples = 1000, w0 = c(0,0)) {
  
}
```

## Suppose we have $g(w_1, w_2=100(w_2-w_1^2)^2+(w_1-w_2)^2$

## (b) Draw the plots of cost_history to showcase your function with $w_0=[-3 -3]^T$.
```{r}

```

## (c) Repeat (b) with a steplength diminishing rule.
```{r}

```

## (d) Discuss your results of (b) and (c)
```{r}

```

\newpage

# (2) Zero-order Coordinate Search

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

## (d) 
```{r}

```

\newpage

# (3) First-order Coordinate Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

\newpage

# (4) Convex Function Properties

## Prove that:

## (1) The sum of convex functions is always convex
```{r}

```

## (2) The maximum of convex functions is convex
```{r}

```

\newpage

# (5) Gradient Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

## (d) 
```{r}

```
