---
title: "Stats 102B Homework 2"
author: "Christian Jordan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r message = FALSE, warning = FALSE, echo = FALSE}

```

# (1) Two-Dimensional Random Search

## (a) random_search()
```{r}
random_search <- function(g, alpha_choice = 1, max_its = 100, num_samples = 1000, w0 = c(0,0)) {
  
  # Initialize weight history
  weight_history <- matrix(w0, nrow = 2)

  # Initialize cost history 
  cost_history <- g(w0[1], w0[2])
  
  its <- 1
  
  while(its <= max_its) {
    current_weights <- weight_history[,its]
    
    weights <- matrix(numeric(0), nrow = 2)
    costs <- numeric(0)
    
    # Generate random directions
    for(i in 1:num_samples) {
      # Random number generation
      direction <- runif(min = -1, max = 1, n = 2)
      direction <- direction / sqrt(sum(direction^2))
      
      # Save random weight update and cost
      weights <- cbind(weights, current_weights + alpha_choice * direction)
      costs <- c(costs, g(weights[1, i], weights[2, i]))
    }
    
    # Save best weights and cost found
    best_weight <- weights[,which(costs == min(costs))]
    best_cost <- g(best_weight[1], best_weight[2])
    
    weight_history <- cbind(weight_history, best_weight)
    cost_history <- c(cost_history, best_cost)
    
    its <- its + 1
  }
  
  list(weight_history, cost_history)
}
```

## Suppose we have $g(w_1, w_2=100(w_2-w_1^2)^2+(w_1-w_2)^2$

```{r}
g <- function(w1, w2) {
  100 * (w2 - w1^2)^2 + (w1 - w2)^2
}
```

## (b) Draw the plots of cost_history to showcase your function with $w_0=[-3 -3]^T$.
```{r}
random_search(g, w0 = c(-3, -3), alpha_choice = 5)
```

## (c) Repeat (b) with a steplength diminishing rule.
```{r}

```

## (d) Discuss your results of (b) and (c)
```{r}

```

\newpage

# (2) Zero-order Coordinate Search

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

## (d) 
```{r}

```

\newpage

# (3) First-order Coordinate Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

\newpage

# (4) Convex Function Properties

## Prove that:

## (1) The sum of convex functions is always convex
```{r}

```

## (2) The maximum of convex functions is convex
```{r}

```

\newpage

# (5) Gradient Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

## (d) 
```{r}

```
