---
title: "Stats 102B Homework 2"
author: "Christian Jordan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r message = FALSE, warning = FALSE, echo = FALSE}

```

# (1) Two-Dimensional Random Search

## (a) random_search()
```{r}
random_search <- function(g, alpha_choice = 1, max_its = 100, num_samples = 1000, w0 = c(0,0)) {
  
  # Initialize weight history
  weight_history <- matrix(w0, nrow = 2)

  # Initialize cost history 
  cost_history <- g(w0[1], w0[2])
  
  its <- 1
  
  while(its <= max_its) {
    current_weights <- weight_history[,its]
    
    weights <- matrix(numeric(0), nrow = 2)
    costs <- numeric(0)
    
    # Generate random directions
    for(i in 1:num_samples) {
      # Random number generation
      direction <- runif(min = -1, max = 1, n = 2)
      direction <- direction / sqrt(sum(direction^2))
      
      # Save random weight update and cost
      if(alpha_choice != "diminishing") {
        weights <- cbind(weights, current_weights + alpha_choice * direction)
        costs <- c(costs, g(weights[1, i], weights[2, i]))
      } else {
        weights <- cbind(weights, current_weights + (1/its) * direction)
        costs <- c(costs, g(weights[1, i], weights[2, i]))
      }
    }
    
    # Save best weights and cost found
    best_index <- which.min(costs)
    best_weight <- weights[, best_index]
    best_cost <- costs[best_index]
    
    weight_history <- cbind(weight_history, unname(best_weight))
    cost_history <- c(cost_history, best_cost)
    
    its <- its + 1
  }
  
  list(weight_history, cost_history)
}
```

## Suppose we have $g(w_1, w_2=100(w_2-w_1^2)^2+(w_1-w_2)^2$

```{r}
g <- function(w1, w2) {
  100 * (w2 - w1^2)^2 + (w1 - w2)^2
}
```

## (b) Draw the plots of cost_history to showcase your function with $w_0=[-3 -3]^T$.
```{r}
cost_hist <- random_search(g, w0 = c(-3, -3))[[2]]
plot(cost_hist)
c(Minima = tail(cost_hist, n = 1))
```

## (c) Repeat (b) with a steplength diminishing rule.
```{r}
cost_hist_d <- random_search(g, w0 = c(-3, -3), alpha_choice = "diminishing")[[2]]
plot(cost_hist_d)
c(Minima = tail(cost_hist, n = 1))
```

## (d) Discuss your results of (b) and (c)

From the plots, we can see that the fixed step length method used in part (b) converges to the estimated optimum sooner (less than 10 iterations) compared to the diminishing-step method used in part (c) which doesn't converge to a final value until after around 60 iterations. However, we can from the code below that the fixed step length method settled at weights resulting in a cost function value of 0.016 while the diminishing step length method settled at weights resulting in a value of 1.01*10^-16. So while the diminishing method took longer to converge, it converged to a better minima for the function. This is because it is less likely to overshoot the actual minima of the function as it gets closer and closer to it more slowly.

```{r}
cost_hist[length(cost_hist)]
cost_hist_d[length(cost_hist_d)]
```

\newpage

# (2) Zero-order Coordinate Search

## (a) 
```{r}
# Pseudocode:
# Function {
# Initialize weight and cost histories
# 
# For iterations 1 to max_its or until tol reached {
#   If odd iteration -> optimize w1
#       Calculate step in + and - w1 directions
#       Save the direction that produces best descent (or neither if no descent)
#       Update cost weight and cost history
#   else, optimize w2
#       Calculate step in + and - w2 directions
#       Save the direction that produces best descent (or neither if no descent)
#       Update cost weight and cost history
# }
# 
# Return final weight and cost
# }
```

```{r}
coordinate_descent_0 <- function(g, alpha_choice = 1, max_its = 100, w0 = c(0,0), tol = 10^-5) {
  
  # Initialize weight history
  weight_history <- matrix(w0, nrow = 2, ncol = 1)

  # Initialize cost history 
  cost_history <- g(w0[1], w0[2])
  
  
  its <- 1
  no_improvement <- 0
  
  while(its <= max_its) {
      
    current_weights <- unname(weight_history[,its])
    current_cost <- cost_history[its]
    
    # Exit if tolerance is met
    if(its > 2 && abs(current_cost - cost_history[its-1]) <= tol && no_improvement >= 2) {
      break
    }
    
      if(alpha_choice == "diminishing") {
        d <- 1/its
      } else {
        d <- alpha_choice
      }
    
    # Alternate coordinate to descend
    if(its %% 2 == 1) {
      weight_plus <- current_weights + c(d,0)
      weight_minus <- current_weights - c(d,0)
    } else {
      weight_plus <- current_weights + c(0, d)
      weight_minus <- current_weights - c(0, d)
    }
    
    cost_plus <- g(weight_plus[1], weight_plus[2])
    cost_minus <- g(weight_minus[1], weight_minus[2])
    
    if(current_cost <= cost_plus && current_cost <= cost_minus) { # Other options are worse
      weight_history <- cbind(weight_history, unname(current_weights))
      cost_history <- c(cost_history, current_cost)
      no_improvement <- no_improvement + 1
    } else {
      if(cost_plus <= cost_minus) { # + direction is better
        weight_history <- cbind(weight_history, unname(weight_plus))
        cost_history <- c(cost_history, cost_plus)
        no_improvement <- 0
      } else { # - direction is better
        weight_history <- cbind(weight_history, unname(weight_minus))
        cost_history <- c(cost_history, cost_minus)
        no_improvement <- 0
      }
    }
  
    its <- its + 1
  }
  
  list(weight_history, cost_history)
}
```

## Suppose we have $g(w_1, w_2=$

```{r}
g2 <- function(w1, w2) {
  w1^2 + w2^2 + 2
}
```

## (b) 
```{r}
# Minimize function
results <- coordinate_descent_0(g2, w0=c(4,4))
weights <- results[1]
costs <- results[2]

# Contour plot
w1_vals <- seq(-2, 5, length.out = 100)
w2_vals <- seq(-2, 5, length.out = 100)
z <- outer(w1_vals, w2_vals, g2)

contour(w1_vals, w2_vals, z,
        xlab = "w1",
        ylab = "w2",
        main = "Contour plot of g(w1, w2)")

# Plot weight values
points(weights[[1]][1,], weights[[1]][2,], col = "red")

# Connect weight values
lines(weights[[1]][1,], weights[[1]][2,],
      col = "red", lwd = 2)

# Plot cost label
text(weights[[1]][1,] - 0.15, weights[[1]][2,]-0.1,
     labels = costs[[1]],
     pos = 3,      
     cex = 0.8,
     col = "blue")
```

## (c) 
```{r}
# Minimize function
results <- coordinate_descent_0(g2, w0=c(4,4), alpha_choice = "diminishing")

weights <- results[1]
costs <- results[2]

# Contour plot
w1_vals <- seq(-2, 5, length.out = 100)
w2_vals <- seq(-2, 5, length.out = 100)
z <- outer(w1_vals, w2_vals, g2)

contour(w1_vals, w2_vals, z,
        xlab = "w1",
        ylab = "w2",
        main = "Contour plot of g(w1, w2)")

# Plot weight values
points(weights[[1]][1,], weights[[1]][2,], col = "red")

# Connect weight values
lines(weights[[1]][1,], weights[[1]][2,],
      col = "red", lwd = 2)

# Plot cost label
text(weights[[1]][1,] - 0.15, weights[[1]][2,]-0.1,
     labels = round(costs[[1]], digits = 2),
     pos = 3,      
     cex = 0.8,
     col = "blue")
```

## (d) 

## Given $g(w_1, w_2=0.26(w_1^2+w_2^2)-0.48w_1w_2$

```{r}
g3 <- function(w1, w2) {
  0.26 * (w1^2 + w2^2) - 0.48 * w1 * w2
}
```

```{r}
coordinate_descent_0(g3, w0=c(4,4), alpha_choice = 0.1)[[2]]

```

\newpage

# (3) First-order Coordinate Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

\newpage

# (4) Convex Function Properties

## Prove that:

## (1) The sum of convex functions is always convex
```{r}

```

## (2) The maximum of convex functions is convex
```{r}

```

\newpage

# (5) Gradient Descent

## (a) 
```{r}

```

## (b) 
```{r}

```

## (c) 
```{r}

```

## (d) 
```{r}

```
