---
title: "Stats 102B Homework 1"
author: "Christian Jordan"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r message = FALSE, warning = FALSE, echo = FALSE}
library(pracma)
```

# (1) Angle Between Two Vectors

## (a) get_angle() Counterclockwise angle between two vectors
```{r}
get_angle = function(x, y) {
  # Calculates the (counterclockwise) angle between two 2-dimensional vectors
  
  # Calculate angle 
  radians <- acos(dot(x,y) / (Norm(x) * Norm(y)))
  degrees <- radians * (180/pi)
  
  # Make angle counterclockwise
  cross_prod <- x[1]*y[2] - x[2]*y[1]

  if(cross_prod > 0) {
    degrees <- 360 - degrees
  }
  
  degrees
}
```

## (b) Vector pairs giving 17, 110, and 270 degrees
```{r}
get_angle(c(cos(17 * (pi/180)),sin(17 * (pi/180))), c(1,0))
get_angle(c(cos(110 * (pi/180)),sin(110 * (pi/180))), c(1,0))
get_angle(c(cos(270 * (pi/180)),sin(270 * (pi/180))), c(1,0))
```
\newpage

# (2) Matrix Inverse Properties

## Given that $A^{-1}$ and $B^{-1}$ exist:

## (a) Show that $(A^T)^{-1}=(A^{-1})^T$

$$AA^{-1}=I$$
$$(AA^{-1})^T=I^T$$
$$A^T(A^{-1})^T=I$$
$$(A^T)^{-1}A^T(A^{-1})^T=(A^T)^{-1}I$$
$$(A^{-1})^T=(A^T)^{-1}$$

## (b) Show that $(AB)^{-1}=B^{-1}A^{-1}$



$$I=AA^{-1}$$
$$I=AIA^{-1}$$
$$I=A(BB^{-1})A^{-1}$$
$$I=(AB)(B^{-1}A^{-1})$$
$$(AB)(AB)^{-1}=(AB)(B^{-1}A^{-1})$$
$$(AB)^{-1}(AB)(AB)^{-1}=(AB)^{-1}(AB)(B^{-1}A^{-1})$$

$$(AB)^{-1}=B^{-1}A^{-1}$$

# (3) Eigenvalue Property

## $A$ is $kxk$ matrix and $P$ is $kxk$ orthogonal matrix. Show that $P^TAP and $A$ have the same eigenvalues$

# (4) Ellipses

# (5) Orthogonal Projection

# (6) Modelling
